Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2019-01-24 RNARedPrint.tex
Fix app refs (partially)

Algorithm~\ref{alg:pf} computes the partition function by passing
messages along the directed edges $u\to v$ (which point from 
child $u$ to its parent $v$). Each message $m$ has the form of a contribution, i.e. it takes a partial sequence, depends
on the positions $\dep(m)\subseteq \X$, and yields a partition function
in $\real$. The message from $u$ to $v$ represents the
partition functions of the subtree of $u$ for all possible partial
sequences in $\partseqs(\separator{u}{v})$. Induction over $T$ lets us show
the correctness of the algorithm
(Supp. Mat.~\ref{appsec:correctness}).  After running
Alg.~\ref{alg:pf}, multiplying the 0-ary messages sent to the root $r$
yields the total partition function (i.e.~due to proper encoding the partition function of our design problem) through
\begin{math}
  \prod_{(u\to{}r)\in T} \evalfor{\Message{u}{r}}{\varnothing}.
\end{math}

@@

%The complexity of the proposed sampling algorithm depends on the treewidth of the dependency graph $G_\F$.
% The treewidth $w$ of a cluster tree $(T,\chi,\phi)$, $T=(V,E)$, is $\max_{v\in V} |\chi(v)| - 1$, i.e.~the maximum number of positions in any of its clusters minus 1.
Let $s$ denote the maximum size of any separator set $\separator{u}{v}$ and $D$ denote the maximum size of
$\difference{u}{v}$ over $(u,v)\in E$.  In the absence
of specific optimizations, running Alg.~\ref{alg:pf} requires
$\mathcal{O}((|\F|+|V|)\cdot 4^{w+1})$ time and
$\mathcal{O}(|V|\cdot4^s)$ space
(Supp. Mat.~\ref{appsec:algcomplexity}); Alg.~\ref{alg:sampling} would
require $\mathcal{O}((|\F|+|V|)\cdot 4^D)$ per sample on arbitrary
tree decompositions
(Supp. Mat.~\ref{appsec:algcomplexity}). W.l.o.g.\ we assume that
$D=1$; note that tree decompositions can generally be transformed,
such that $\difference{u}{v}\leq 1$.
%
Moreover, the size of $\F$ is linearly bounded: for $k$
input structures for sequences of length $n$, the energy function is
expressed by $\mathcal{O}(n\,k)$ functions. Finally, the number of cluster
tree nodes is in $O(n)$, such that $|\F|+|V| \in \mathcal{O}(n\,k)$.

@@

By this theorem, the complexity is polynomial for fixed value of $w$, and Boltzmann sampling in our setting is thus fixed parameter tractable (FPT) in the treewidth.
%
As shown in Supp. Mat.~\ref{sec:improvedComplexity}, the complexity of the precomputation can be further improved to
$\mathcal{O}(n\,k\,2^{w+1}\,2^{c})$, where $c$ is the maximum number of connected components represented in a node of the tree decomposition ($c\le w+1$).

Note that in this complexity analysis, we do not include time and space for computing the
tree decomposition itself, since we observed that the computation time
of tree decomposition (\Software{GreedyFillIn}, implemented in
\Software{LibTW}~by \citet{Dijk2006}) for multi-target sampling is
negligible compared to Alg.~\ref{alg:pf}
(Supp. Mat.~\ref{appsec:treedecomp} and
\ref{appsec:dependency-cliques}).

@@

\paragraph{Approximating Turner Energy using Simpler Energy Models.}
To capture the realistic Turner model $\EnergyTurner$ more efficiently, we exploit the tight correlation between
$\EnergyTurner$ and the fitted stacking model $\EnergyStacking$
(Supp. Fig~\ref{fig:training-cor}). More precisely, we observed a
structure-specific affine dependency between the Turner and stacking
energy models, so that
$\EnergyTurner(S;R) \approx \gamma\cdot \EnergyStacking(S;R) + \delta$
for any structure $R$ and sequence $S$. We inferred the
$(\gamma,\delta)$ parameters from a set of sequences generated with
homogenous weights $w=e^{\beta}$, tuning only \GCb\% to a
predetermined value.  Finally, we adjusted the targeted energies
within our stacking model to
$\EnergyStacking^{\star} = (\EnergyTurner^{\star}- \delta)/\gamma$ in
order to reach, on average, the targeted energy
$\EnergyTurner^{\star}$ in the Turner model.

@@

One then needs to \Def{learn a weights vector} $\pmb{\pi}$ such that, on average, the targeted energies are achieved by a random sequences in the weighted distribution. In other words,  $\mathbb{E}(F_\ell(S)\mid \pmb{\pi})=\TargetE_\ell$,  $\forall\ell\in[1,k]$ and, analogously, the expectation of $F_0(S)$ is the targeted GC content.
The expected value of $F_\ell$ is always decreasing for increasing weights $\pi_\ell$ (see Supp. Mat.~\ref{sec:weight_derivatives}). More generally, computing a suitable parameter vector $\pmb{\pi}$ can be restated as a convex optimization problem, and be efficiently solved using a wide array of methods~\citep{Denise2010,Bendkowski2017}.

@@

We used our Boltzmann sampling strategy (Algorithms
\ref{alg:pf} and \ref{alg:sampling}), to sample valid sequences for
given target structures and weights $\pi_1,\dots,\pi_k$.  Moreover, we
used multi-dimensional Boltzmann sampling to target specific energies and
\GCb\%.  Our tool \ourprog{} evaluates energies according to the
stacking energy model $\EnergyStacking$, whose parameters were fitted
to best approximate Turner energies. As well, we implemented and
fitted a base pair energy model for \ourprog{}, which was not studied
for its targeting performance (both models:
Supp. Mat.~\ref{appsec:modelparameters}).
%
%\paragraph{Implementation of multi-dimensional Boltzmann sampling}
%As suggested in Section~\ref{sec:multiBoltzmann}, we heuristically optimize weights to target specific energies and \GCb\% in an iterative procedure (based on the core sampling algorithm).
%Specifically, energy weights are initialized as $w_\ell=e^{\frac{1}{RT}}$; the gc weight as $w_{gc}=1$. For each iteration, $\psi\cdot n$ sequences are sampled with the current weights. Then, we determine the mean energies $E(S;R_\ell)$ and the mean GC content over the sampled sequences. For the next iteration, the weights are updated by the rules $w_\ell = w_\ell \gamma^{mean(E(S;R_\ell))-E_\ell}$ and $w_{gc} = w_{gc}\frac{GC_{target}}{mean(GC)}$. During this procedure, we collect   in a set $P$ all the sequences exhibiting energies within $[E_1-\delta,E_1+\delta]\times [E_2-\delta,E_2+\delta]\ldots$ and and GC-content within $[GC_{target}-\idelta',GC_{target}+\delta']$.
%If $|P| > n$ or an maximum amount of iterations is reached, the sequences in $P$ are returned.
%

Fig.~\ref{fig:energydist-pk} illustrates how well complex realistic energy models can be approximated based on simpler, but better tractable ones. For
the two target structures of Fig.~\ref{fig:energydist-pk}B,
Fig.~\ref{fig:energydist-pk}A shows the good fit between realistic
energies in the full-fledged Dirks and Pierce energy model for
pseudoknots (D\&P model) and energies in the stacking energy
model. For the shown fits we sampled $n=1\,000$ sequences, targeting a
\GCb\% of $60\%$.
%
For an example instance of the \texttt{Modena} benchmark with two
pseudoknotted target structures, Fig.~\ref{fig:energydist-pk}B shows
the Turner energy distributions of the single structures as they
result from sampling with different weight parameters. The figure
illustrates how our multidimensional Boltzmann sampling strategy can,
to a large extent, independently shift the Turner energies of sampled
sequences towards prescribed targets. See Supp. Mat.,
Fig.~\ref{fig:energydist} for a further example with three
pseudoknot-free target structures.}

@@

The results, shown in Fig.~\ref{fig:benchmark-results}, reveal that
Boltzmann-sampled sequences outperform uniform seeds on every dataset,
leading to average improvements in \Obj{} values ranging from $7.26$
({\sf LE80}) to $16.05$ ({\sf 2str}) units. Remarkably, this
improvement is be observed for both of the terms in \Obj{} (see
Supp. Mat.~\ref{sec:improvements}).  This means that \ourprog{}
produces sequences whose targets are both substantially closer to
the sequence ensemble/minimal free energy, and more similarly stable across targets.  In fact, for
every sequence in our benchmark, consisting of 332 sets of target
structures, we observed better \Obj{} for Boltzmann sampling than for
uniform sampling (see Supp. Mat.~\ref{sec:validity}).  Notably,
\ourprog{} performs equally well in the presence of pseudoknots;
the difficulty rather lies in the computation of the \Obj{} function, since
free-energy minimization is costly in the presence of
pseudoknots~\citep{Sheikh2012} and good implementations are scarce.

%
Moreover, for all instances as well, the superiority of Boltzmann designs
remains unchallenged even after local optimizations, as shown by
Fig.~\ref{fig:benchmark-results} and Supp. Mat.~\ref{sec:improvements}
and \ref{sec:validity}. This observation is consistent with a
superior quality of the starting point for the greedy walk, probably leading
to better local minima of the \Obj{} function. However, it should be noted
that the greedy walk is based on the uniform sampling of
\RNAblueprint{}, and thus can be expected to partially level the
advantages of Boltzmann sampling. In future work, we hope to improve
this aspect by exploiting Boltzmann sampling during the
optimization run.

>>>>

Algorithm~\ref{alg:pf} computes the partition function by passing
messages along the directed edges $u\to v$ (which point from 
child $u$ to its parent $v$). Each message $m$ has the form of a contribution, i.e. it takes a partial sequence, depends
on the positions $\dep(m)\subseteq \X$, and yields a partition function
in $\real$. The message from $u$ to $v$ represents the
partition functions of the subtree of $u$ for all possible partial
sequences in $\partseqs(\separator{u}{v})$. Induction over $T$ lets us show
the correctness of the algorithm
(Supp.\ Mat.~\ref{appsec:correctness}).  After running
Alg.~\ref{alg:pf}, multiplying the 0-ary messages sent to the root $r$
yields the total partition function (i.e.~due to proper encoding the partition function of our design problem) through
\begin{math}
  \prod_{(u\to{}r)\in T} \evalfor{\Message{u}{r}}{\varnothing}.
\end{math}

@@

%The complexity of the proposed sampling algorithm depends on the treewidth of the dependency graph $G_\F$.
% The treewidth $w$ of a cluster tree $(T,\chi,\phi)$, $T=(V,E)$, is $\max_{v\in V} |\chi(v)| - 1$, i.e.~the maximum number of positions in any of its clusters minus 1.
Let $s$ denote the maximum size of any separator set $\separator{u}{v}$ and $D$ denote the maximum size of
$\difference{u}{v}$ over $(u,v)\in E$.  In the absence
of specific optimizations, running Alg.~\ref{alg:pf} requires
$\mathcal{O}((|\F|+|V|)\cdot 4^{w+1})$ time and
$\mathcal{O}(|V|\cdot4^s)$ space; Alg.~\ref{alg:sampling} would
require $\mathcal{O}((|\F|+|V|)\cdot 4^D)$ per sample on arbitrary
tree decompositions
(Supp.\ Mat.~\ref{appsec:algcomplexity}). W.l.o.g.\ we assume that
$D=1$; note that tree decompositions can generally be transformed,
such that $\difference{u}{v}\leq 1$.
%
Moreover, the size of $\F$ is linearly bounded: for $k$
input structures for sequences of length $n$, the energy function is
expressed by $\mathcal{O}(n\,k)$ functions. Finally, the number of cluster
tree nodes is in $O(n)$, such that $|\F|+|V| \in \mathcal{O}(n\,k)$.

@@

By this theorem, the complexity is polynomial for fixed value of $w$, and Boltzmann sampling in our setting is thus fixed parameter tractable (FPT) in the treewidth.
%
The complexity of the precomputation can be further improved to
$\mathcal{O}(n\,k\,2^{w+1}\,2^{c})$, where $c$ ($c\le w+1$) is the maximum number of connected components represented in a node of the tree decomposition (Supp.\ Mat.~\ref{appsec:improvedComplexity}).

Note that in this complexity analysis, we do not include time and space for computing the
tree decomposition itself, since we observed that the computation time
of tree decomposition (\Software{GreedyFillIn}, implemented in
\Software{LibTW}~by \citet{Dijk2006}) for multi-target sampling is
negligible compared to Alg.~\ref{alg:pf}
(Supp.\ Mat.~\ref{appsec:treedecomp} and
\ref{appsec:dependency-cliques}).

@@

\paragraph{Approximating Turner Energy using Simpler Energy Models.}
To capture the realistic Turner model $\EnergyTurner$ more efficiently, we exploit the tight correlation between
$\EnergyTurner$ and the fitted stacking model $\EnergyStacking$
(Supp.\ Mat.~\ref{appsec:modelparameters}). More precisely, we observed a
structure-specific affine dependency between the Turner and stacking
energy models, so that
$\EnergyTurner(S;R) \approx \gamma\cdot \EnergyStacking(S;R) + \delta$
for any structure $R$ and sequence $S$. We inferred the
$(\gamma,\delta)$ parameters from a set of sequences generated with
homogenous weights $w=e^{\beta}$, tuning only \GCb\% to a
predetermined value.  Finally, we adjusted the targeted energies
within our stacking model to
$\EnergyStacking^{\star} = (\EnergyTurner^{\star}- \delta)/\gamma$ in
order to reach, on average, the targeted energy
$\EnergyTurner^{\star}$ in the Turner model.

@@

One then needs to \Def{learn a weights vector} $\pmb{\pi}$ such that, on average, the targeted energies are achieved by a random sequences in the weighted distribution. In other words,  $\mathbb{E}(F_\ell(S)\mid \pmb{\pi})=\TargetE_\ell$,  $\forall\ell\in[1,k]$ and, analogously, the expectation of $F_0(S)$ is the targeted GC content.
The expected value of $F_\ell$ is always decreasing for increasing weights $\pi_\ell$ (see Supp.\ Mat.~\ref{appsec:weight_derivatives}). More generally, computing a suitable parameter vector $\pmb{\pi}$ can be restated as a convex optimization problem, and be efficiently solved using a wide array of methods~\citep{Denise2010,Bendkowski2017}.

@@

We used our Boltzmann sampling strategy (Algorithms
\ref{alg:pf} and \ref{alg:sampling}), to sample valid sequences for
given target structures and weights $\pi_1,\dots,\pi_k$.  Moreover, we
used multi-dimensional Boltzmann sampling to target specific energies and
\GCb\%.  Our tool \ourprog{} evaluates energies according to the
stacking energy model $\EnergyStacking$, whose parameters were fitted
to best approximate Turner energies. As well, we implemented and
fitted a base pair energy model for \ourprog{}, which was not studied
for its targeting performance (both models:
Supp.\ Mat.~\ref{appsec:modelparameters}).
%
%\paragraph{Implementation of multi-dimensional Boltzmann sampling}
%As suggested in Section~\ref{sec:multiBoltzmann}, we heuristically optimize weights to target specific energies and \GCb\% in an iterative procedure (based on the core sampling algorithm).
%Specifically, energy weights are initialized as $w_\ell=e^{\frac{1}{RT}}$; the gc weight as $w_{gc}=1$. For each iteration, $\psi\cdot n$ sequences are sampled with the current weights. Then, we determine the mean energies $E(S;R_\ell)$ and the mean GC content over the sampled sequences. For the next iteration, the weights are updated by the rules $w_\ell = w_\ell \gamma^{mean(E(S;R_\ell))-E_\ell}$ and $w_{gc} = w_{gc}\frac{GC_{target}}{mean(GC)}$. During this procedure, we collect   in a set $P$ all the sequences exhibiting energies within $[E_1-\delta,E_1+\delta]\times [E_2-\delta,E_2+\delta]\ldots$ and and GC-content within $[GC_{target}-\idelta',GC_{target}+\delta']$.
%If $|P| > n$ or an maximum amount of iterations is reached, the sequences in $P$ are returned.
%

Fig.~\ref{fig:energydist-pk} illustrates how well complex realistic energy models can be approximated based on simpler, but better tractable ones. For
the two target structures of Fig.~\ref{fig:energydist-pk}B,
Fig.~\ref{fig:energydist-pk}A shows the good fit between realistic
energies in the full-fledged Dirks and Pierce energy model for
pseudoknots (D\&P model) and energies in the stacking energy
model. For the shown fits we sampled $n=1\,000$ sequences, targeting a
\GCb\% of $60\%$.
%
For an example instance of the \texttt{Modena} benchmark with two
pseudoknotted target structures, Fig.~\ref{fig:energydist-pk}B shows
the Turner energy distributions of the single structures as they
result from sampling with different weight parameters. The figure
illustrates how our multidimensional Boltzmann sampling strategy can,
to a large extent, independently shift the Turner energies of sampled
sequences towards prescribed targets. See Supp.\ Mat.~\ref{appsec:illustrating-mdbs} for a further example with three
pseudoknot-free target structures.}

@@

The results, shown in Fig.~\ref{fig:benchmark-results}, reveal that
Boltzmann-sampled sequences outperform uniform seeds on every dataset,
leading to average improvements in \Obj{} values ranging from $7.26$
({\sf LE80}) to $16.05$ ({\sf 2str}) units. Remarkably, this
improvement is be observed for both of the terms in \Obj{} (see
Supp. Mat.~\ref{appsec:contributions-multidefect}).  This means that \ourprog{}
produces sequences whose targets are both substantially closer to
the sequence ensemble/minimal free energy, and more similarly stable across targets.  In fact, for
every sequence in our benchmark, consisting of 332 sets of target
structures, we observed better \Obj{} for Boltzmann sampling than for
uniform sampling (see Supp. Mat.~\ref{appsec:detailed-benchmark-results}).  Notably,
\ourprog{} performs equally well in the presence of pseudoknots;
the difficulty rather lies in the computation of the \Obj{} function, since
free-energy minimization is costly in the presence of
pseudoknots~\citep{Sheikh2012} and good implementations are scarce.

%
Moreover, for all instances as well, the superiority of Boltzmann designs
remains unchallenged even after local optimizations, as shown by
Fig.~\ref{fig:benchmark-results} and Supp. Mat.~\ref{appsec:contributions-multidefect}
and \ref{appsec:detailed-benchmark-results}. This observation is consistent with a
superior quality of the starting point for the greedy walk, probably leading
to better local minima of the \Obj{} function. However, it should be noted
that the greedy walk is based on the uniform sampling of
\RNAblueprint{}, and thus can be expected to partially level the
advantages of Boltzmann sampling. In future work, we hope to improve
this aspect by exploiting Boltzmann sampling during the
optimization run.
